Some references on EDCs, glucocorticosteroids, and neurodevelopment

### 2023-12-31 docs/bib_edc_cort_neuro.qmd
Finished doc with references.

---
title: "Some references on EDCs, corticosteroids, and neurodevelopment"

@@

- Title: **.
- Exposures: .
- Medium exposure assessment: .
- Timing exposure assessment: .
- Outcomes: .
- Medium outcome assessment: .
- Timing outcome assessment: .
- Results:
    - .
- Sex-specific results: .

@@

- Title: **.
- Exposures: .
- Medium exposure assessment: .
- Timing exposure assessment: .
- Outcomes: .
- Medium outcome assessment: .
- Timing outcome assessment: .
- Results:
    - .
- Sex-specific results: .

@@

- Title: **.
- Exposures: .
- Medium exposure assessment: .
- Timing exposure assessment: .
- Outcomes: .
- Medium outcome assessment: .
- Timing outcome assessment: .
- Results:
    - .
- Sex-specific results: .

@@

TODO:

- Gascon et al. 2015
- Oh et al. 2023
- Vilmand et al. 2023

@@

- Title: **.
- Exposures: .
- Medium exposure assessment: .
- Timing exposure assessment: .
- Outcomes: .
- Medium outcome assessment: .
- Timing outcome assessment: .
- Results:
    - .
- Sex-specific results: .

@@

- Title: **.
- Exposures: .
- Medium exposure assessment: .
- Timing exposure assessment: .
- Outcomes: .
- Medium outcome assessment: .
- Timing outcome assessment: .
- Results:
    - .
- Sex-specific results: .

TODO:

- Kim et al. 2018
- Sun et al. 2018

@@

>>>>

---
title: "Some references on EDCs, glucocorticosteroids, and neurodevelopment"

@@

@@

## @OhKimKannan:2023

- Title: *Early childhood exposure to environmental phenols and parabens, phthalates, organophosphate pesticides, and trace elements in association with attention de cit hyperactivity disorder (ADHD) symptoms in the CHARGE study*.
- Exposures: phenols/parabens, phthalates, OP pesticides, trace elements.
- Medium exposure assessment: spot urine samples.
- Timing exposure assessment: ages 2-5.
- Outcomes: ADHD symptoms.
- Timing outcome assessment: ages 2-5.
- Results:
    - Positive associations of MECPP and MEHHP with Hyperactivity subscale scores and hyperactivity/impulsivity subdomain scores.
    - Positive associations of MNBP with hyperactivity/impulsivity subdomain scores.
    - Inverse associations of TCS with Hyperactivity subscale scores.
    - Inverse associations of TCS with inattention subdomain scores.
    - Positive associations of mixture of phthalates with scores of the ABC Hyperactivity subscale and hyperactivity/impulsivity subdomain.
      - Drivers by subdomain, hyperactivity/impulsivity subdomain: MHPP, MEHHP, and MBZP.
    - Positive associations of mixture of phenols/parabens with inattention subdomain score.
      - Drivers by subdomain, inattention subdomain: MEPB, PRPB, and OHMEP.
- Sex-specific results: association of mixture of phthalates with hyperactivity-related subscale and subdomain significant in males only.

@@

## @VilmandBeckBilenberg:2023

- Title: *Prenatal and current phthalate exposure and cognitive development in 7-year-old children from the Odense child cohort*.
- Exposures: phthalates metabolites.
- Medium exposure assessment: spot urine samples.
- Timing exposure assessment: 7 years.
- Outcomes: IQ (subtests from Wechsler Intelligence Scale for Children).
- Timing outcome assessment: 7 years.
- Results:
    - Inverse associations of MCPP, sum of DnHxPm, sum of DiDPm, and sum of DiNPm with IQ points.
- Sex-specific results: no evidence of effect modification by sex.

@@

@@

## @KimLeeMoon:2018

- Title: *Association of phthalate exposures with urinary free cortisol and 8-hydroxy-2'-deoxyguanosine in early childhood*.
- Exposures: four phthalate metabolites.
- Medium exposure assessment: spot urine samples.
- Timing exposure assessment: prenatal and at ages 0, 3, 9, 12, 15 months.
- Outcomes: free cortisol.
- Medium outcome assessment: spot urine samples.
- Timing outcome assessment: prenatal and at ages 0, 3, 9, 12, 15 months.
- Results:
    - Positive associations between MEHHP, MEOHP, MiBP, and MnBP, with free cortisol.
- Sex-specific results: no evidence of modification by sex.

@@

## @SunLiJin:2018

- Title: *Associations between repeated measures of maternal urinary phthalate metabolites during pregnancy and cord blood glucocorticoids*.
- Exposures: phthalate metabolites.
- Medium exposure assessment: spot urine samples.
- Timing exposure assessment: 14, 24, and 36 weeks of gestation.
- Outcomes: cortisol and cortisone.
- Medium outcome assessment: cord blood.
- Timing outcome assessment: 14, 24, and 36 weeks of gestation.
- Results:
    - Positive association of MEHP 3rd trimester with cortisol.
    - Inverse association of MECPP and MEOHP 3rd trimester with cortisone.
    - Positive association of MBzP 1st trimester with cortisol/cortisone ration.
- Sex-specific results: evidence of modification by sex, but not significant.
    - Females: positive associations with glucocorticoids (MEHP and MBzP 1st trimester, and DEHP 3rd trimester).
    - Males: inverse associations with glucocorticoids (MECPP, MEHHP, and MEOHP 3rd trimester).

@@

## @LupienMcEwenGunnar:2009

- Title: *Effects of stress throughout the lifespan on the brain, behaviour and cognition*.

The HPA axis, which can be activated by stress, is responsible for the production of glucocorticoids. The brain, and its proper functioning, is a potential target, due to the presence of receptors for these hormones.
Glucocorticoids are necessary for brain maturation, although their under- or over-production might interfere with its normal development and ultimately lead to long-term impaired functioning.
